Guarded Recursion in Agda via Sized Types

### 2019-02-01 code-oneclock-paper/latex/Introduction.tex
conclusions and overview

Dependent type theory is an expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda \cite{norell} and Lean.  It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
are required to be productive in the sense that every
finite unfolding can be computed in a finite number of steps \cite{Coquand93}.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, the recursive
call is required to appear directly under
the application of a constructor. A similar syntactic
check is employed in Coq.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
 is a type annotated with the number of unfolding that
elements of this type can undergo \cite{Hughes96}. Sized types have been implemented
in Agda and can be used in
combination with coinductive records to specify
coinductive types \cite{Copatterns,AVW-normalization}.  Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \Ar{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \Ar{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus with guarded recursion.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory with guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aformentioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. In this respect \GTT\ is
closer in spirit to Atkey and McBride's language. The design of \GTT\
also has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

Abel and Vezzosi \cite{AbelV14} also formalized a simple type theory
extended with the later modality in Agda. They focus on operational
properties of the calculus and give a certified proof of strong
normalization. In our work we consider a richer type system and we
formalize its denotational semantics in Agda. Nevertheless, we do not
investigate the dynamic behaviour of our calculus, which we leave to
future work.

%% A detailed account of how to model the characteristic features of
%% \GTT, consisting of the \IC{▻} and
%% \IC{□}  modality together with guarded recursive types, is given in
%% Section \ref{sec:guarded}.

To summarize, the contributions of this paper are twofold:
\begin{enumerate}
  \item We formalize syntax and semantics of a type theory with
    guarded recursion in Agda. This is the first such denotational model developed using an extension of Martin-L\"of intensional
    type theory as the metalanguage, in contrast with the previous
    set-theoretic models of guarded recursion.
\item The interpretation of the characteristic features of guarded
  recursion crucially requires the employment of sized types. This shows that
  guarded recursive types can be reduced to sized types and
  constitutes a stepping stone towards a clear understanding of  the connections
  between different approaches to productive coprogramming.
%	\item We formally prove the syntax is consistent.
\end{enumerate}

@@

>>>>

Dependent type theory is an expressive functional programming
language that underlies the deductive system of proof assistants such
as Coq, Agda, and Lean \cite{barras1997coq,de2015lean,norell}. It is a total language,
meaning that every program definable inside type theory is necessarily
terminating. This is an important requirement that ensures the
consistency of the type system.

Possibly non-terminating computations
and infinite structures such as non-wellfounded trees can be
represented in type theory by extending the type system with
coinductive types. Recursively defined elements of these types
are required to be productive in the sense that every
finite unfolding can be computed in a finite number of steps \cite{Coquand93}.
In Agda's encoding of coinductive types using ``musical notation''
\cite{DA10}, productivity is enforced via a strict obligation: in the
definition of a corecursive function, the recursive
call is required to appear directly under
the application of a constructor. A similar syntactic
check is employed in Coq \cite{DBertotK08}.
This restriction
typically makes programming with coinductive types cumbersome,
which spawned the search for alternative techniques for ensuring the
well-definedness of corecursive definitions.

We focus on two of these techniques in which productivity is encoded
at the level of types: sized types and guarded recursion. A sized type
 is a type annotated with the number of unfolding that
elements of this type can undergo \cite{Hughes96}. Sized types have been implemented
in Agda and can be used in
combination with coinductive records to specify
coinductive types \cite{Copatterns,AVW-normalization}.  Guarded recursion \cite{Nakano}
is a different approach where the type system is enhanced with a
modality, called ``later'' and written \IC{▻}, encoding time delay in
types. The later modality comes with a general fixpoint combinator for
programming with productive recursive functions and allows the
specification of guarded recursive types, \eg the type of guarded
streams satisfying the isomorphism \F{Str} \Ar{A} \F{≃} \Ar{A} \F{×}
\IC{▻} (\F{Str} \Ar{A}). The tail of a guarded stream is available only
one time step ahead from now, as evidenced by the application of the
later modality to \F{Str} \Ar{A}. Coinductive types can be constructed
from guarded recursive types in type theories that additionally
support clock variables and universal quantification over them
\cite{atkey2013productive,BahrGM17}. Currently there is no
implementation of a calculus for guarded recursion.
%%  nor guarded recursion
%% appears as a selectable feature in a proof assistant.

@@

Utilizing techniques for representing ``type theory in type theory''
\cite{AltenkirchK16,Chapman09}, we present an Agda formalization of
the syntax of a simple type theory for guarded recursion. This object
language, which we call \GTT, is a variant of Atkey and McBride's type
system for productive coprogramming \cite{atkey2013productive} in
which the clock context can contain at most one clock. The types of
\GTT\ include the aformentioned \IC{▻} modality and a \IC{□} modality,
a nameless analogue of Atkey and McBride's universal clock
quantification. Clouston \etal \cite{CloustonBGB15} studied a
guarded variant of lambda calculus extended with a \IC{□} operation,
which they call ``constant''.  Our object calculus differs from
theirs in that our judgments are indexed by a clock context, which can
be empty or containing exactly one clock. In this respect \GTT\ is
closer in spirit to Atkey and McBride's language. The design of \GTT\
also has the benefit of allowing a more appealing introduction rule for
\IC{□} than Clouston et al.'s.

@@

%%Abel and Vezzosi \cite{AbelV14} also formalized a simple type theory
%%extended with the later modality in Agda. They focus on operational
%%properties of the calculus and give a certified proof of strong
%%normalization. In our work we consider a richer type system and we
%%formalize its denotational semantics in Agda. Nevertheless, we do not
%%investigate the dynamic behaviour of our calculus, which we leave to
%%future work.

To summarize, the contributions of this paper are twofold:
\begin{enumerate}
  \item We formalize syntax and semantics of a type theory for
    guarded recursion in Agda. This is the first such denotational model developed using an extension of Martin-L\"of intensional
    type theory as the metalanguage, in contrast with the previous
    set-theoretic models of guarded recursion.
\item The interpretation of the characteristic features of guarded
  recursion crucially requires the employment of sized types. This shows that
  guarded recursion can be reduced to sized types and
  constitutes a stepping stone towards a clear understanding of  the connections
  between different approaches to productive coprogramming.
%	\item We formally prove the syntax is consistent.
\end{enumerate}

@@

The material is organized in the following way.
In \Cref{sec:prelim}, we discuss the metatheory and we give an
overview of sized types.
In \Cref{sec:syntax}, we introduce the syntax of \GTT, our object type theory for guarded recursion.
Subsequently, we give a categorical semantics.
In \Cref{sec:presheaf_sem}, we show how to implement presheaves over
sizes and how to model the fragment of
\GTT\ corresponding to simply typed lambda calculus.
In \Cref{sec:guarded}, we discuss the semantics of the guarded
recursive and coinductive features.
In \Cref{sec:consistent}, we describe how to formally interpret the
syntax into the model, which entails the consistency of the syntax.
Finally, in \Cref{sec:concl}, we draw conclusions and suggest  future
directions of work.

%% give the Kripke semantics
%%\GTT.
%%For that, we first define a presheaf model and give the Kripke semantics, and then we show how to interpret the additional type and term formers of guarded recursive type theory.
%%Finally, we conclude that the syntax is consistent in \Cref{sec:consistent}.

### 2019-02-01 code-oneclock-paper/latex/paper.tex
Sec 5.2 until line 429
Added overview of paper
Added a unicode symbol for ▸
conclusions and overview

\newunicodechar{⊤}{$\top$}
\newunicodechar{≡}{$\equiv$}
\newunicodechar{₁}{$_1$}
\newunicodechar{₂}{$_2$}
\newunicodechar{₃}{$_3$}
\newunicodechar{Γ}{$\Gamma$}
\newunicodechar{Δ}{$\Delta$}
\newunicodechar{Σ}{$\Sigma$}
\newunicodechar{α}{$\alpha$}
\newunicodechar{β}{$\beta$}
\newunicodechar{η}{$\eta$}
\newunicodechar{γ}{$\gamma$}
\newunicodechar{ε}{$\varepsilon$}
\newunicodechar{θ}{$\theta$}
\newunicodechar{λ}{$\lambda$}
\newunicodechar{μ}{$\mu$}
\newunicodechar{κ}{$\kappa$}
\newunicodechar{π}{$\pi$}
\newunicodechar{∈}{$\in$}
\newunicodechar{⊞}{$\boxplus$}
\newunicodechar{⊠}{$\boxtimes$}
\newunicodechar{∁}{$\mathbb{C}$}
\newunicodechar{►}{$\blacktriangleright$}
\newunicodechar{▻}{$\vartriangleright$}
\newunicodechar{𝟙}{\textbf{1}}
\newunicodechar{∙}{$\cdot$}
\newunicodechar{∀}{$\forall$}
\newunicodechar{ℓ}{$\ell$}
\newunicodechar{∅}{$\varnothing$}
\newunicodechar{⟶}{$\rightarrow$}
\newunicodechar{□}{$\square$}
\newunicodechar{⇡}{$\uparrow$}
\newunicodechar{⊛}{$\circledast$}
\newunicodechar{⊎}{$\uplus$}
\newunicodechar{⟨}{$\langle$}
\newunicodechar{⟩}{$\rangle$}
\newunicodechar{■}{$\blacksquare$}
\newunicodechar{∎}{$\blacksquare$}
\newunicodechar{⇒}{$\Rightarrow$}
\newunicodechar{⊕}{$\oplus$}
\newunicodechar{⊗}{$\otimes$}
\newunicodechar{∼}{$\sim$}
\newunicodechar{≈}{$\approx$}
\newunicodechar{⊔}{$\sqcup$}
\newunicodechar{⟦}{$\llbracket$}
\newunicodechar{⟧}{$\rrbracket$}
\newunicodechar{⊥}{$\bot$}
\newunicodechar{∞}{$\infty$}
\newunicodechar{⇑}{$\Uparrow$}
\newunicodechar{≃}{$\simeq$}
\newunicodechar{∘}{$\circ$}
% Add more as you need them (shouldn't happen often).

@@

\subsection{Modelling Simple Types}
%\label{sec:kripke}
\input{CloTT/Structure/Contexts}
\input{CloTT/Structure/Types}
\input{CloTT/Structure/Terms}
\input{CloTT/Structure/Subst}
\input{CloTT/Structure/DefinitionalEquality}

@@

\subsection{Guarded Recursive Types}
\input{CloTT/TypeFormers/Mu}

\section{Consistency}
\input{GuardedTT}

\section{Conclusion}
We formally defined a syntax of guarded type theory with explicit substitutions and weakening, and we constructed denotational semantics for it in type theory extended with sized types.
From this, we concluded that this theory is consistent and that guarded recursion can be defined via sized types.

There are several possible ways to improve on this result.
First of all, one could define a dependent version of this type theory extended with multiple clocks.
This can be done in two ways.
Either one extends the object theory with delayed substitutions \cite{bizjak2016guarded} or with ticks \cite{BahrGM17}.
Second of all, with the deep embedding in place, one can further study the meta-theory by defining a formalized type checking and a normalization algorithm.

>>>>

\newunicodechar{⊤}{$\top$}
\newunicodechar{≡}{$\equiv$}
\newunicodechar{₁}{$_1$}
\newunicodechar{₂}{$_2$}
\newunicodechar{₃}{$_3$}
\newunicodechar{Γ}{$\Gamma$}
\newunicodechar{Δ}{$\Delta$}
\newunicodechar{Σ}{$\Sigma$}
\newunicodechar{α}{$\alpha$}
\newunicodechar{β}{$\beta$}
\newunicodechar{η}{$\eta$}
\newunicodechar{γ}{$\gamma$}
\newunicodechar{ε}{$\varepsilon$}
\newunicodechar{θ}{$\theta$}
\newunicodechar{λ}{$\lambda$}
\newunicodechar{μ}{$\mu$}
\newunicodechar{κ}{$\kappa$}
\newunicodechar{π}{$\pi$}
\newunicodechar{∈}{$\in$}
\newunicodechar{⊞}{$\boxplus$}
\newunicodechar{⊠}{$\boxtimes$}
\newunicodechar{∁}{$\mathbb{C}$}
\newunicodechar{▸}{$\blacktriangleright$}
\newunicodechar{►}{$\blacktriangleright$}
\newunicodechar{▻}{$\vartriangleright$}
\newunicodechar{𝟙}{\textbf{1}}
\newunicodechar{∙}{$\cdot$}
\newunicodechar{∀}{$\forall$}
\newunicodechar{ℓ}{$\ell$}
\newunicodechar{∅}{$\varnothing$}
\newunicodechar{⟶}{$\rightarrow$}
\newunicodechar{□}{$\square$}
\newunicodechar{⇡}{$\uparrow$}
\newunicodechar{⊛}{$\circledast$}
\newunicodechar{⊎}{$\uplus$}
\newunicodechar{⟨}{$\langle$}
\newunicodechar{⟩}{$\rangle$}
\newunicodechar{■}{$\blacksquare$}
\newunicodechar{∎}{$\blacksquare$}
\newunicodechar{⇒}{$\Rightarrow$}
\newunicodechar{⊕}{$\oplus$}
\newunicodechar{⊗}{$\otimes$}
\newunicodechar{∼}{$\sim$}
\newunicodechar{≈}{$\approx$}
\newunicodechar{⊔}{$\sqcup$}
\newunicodechar{⟦}{$\llbracket$}
\newunicodechar{⟧}{$\rrbracket$}
\newunicodechar{⊥}{$\bot$}
\newunicodechar{∞}{$\infty$}
\newunicodechar{⇑}{$\Uparrow$}
\newunicodechar{≃}{$\simeq$}
\newunicodechar{∘}{$\circ$}
% Add more as you need them (shouldn't happen often).

@@

\subsection{Modelling Simple Types}
\label{sec:kripke}
\input{CloTT/Structure/Contexts}
\input{CloTT/Structure/Types}
\input{CloTT/Structure/Terms}
\input{CloTT/Structure/Subst}
\input{CloTT/Structure/DefinitionalEquality}

@@

\subsection
{Guarded Recursive Types}
\label{sec:grt}
\input{CloTT/TypeFormers/Mu}

\section{Consistency}
\label{sec:consistent}
\input{GuardedTT}

\section{Conclusions and Future Work}
\label{sec:concl}
We presented a simple type theory for guarded recursion that we called \GTT. We
formalized its syntax and semantics in Agda. Crucially, sized types
were employed for giving an interpretation to the characteristic
features of guarded recursion. From this we conclude that guarded
recursion can be simulated using sized types. We greatly benefited
of the fact that sized types constitute a native feature of Agda, so
we were able to fully develop our theory inside a proof assistent.

This work can be extended is several different directions.

Various type theories for guarded recursion in the literature include
dependent types. Currently, there exist two disciplines for mixing
dependent types with guarded recursion: delayed substitutions
\cite{bizjak2016guarded} and ticks \cite{BahrGM17}. We plan to extend
the syntax and semantics of \GTT\ with dependent types following one
of these two disciplines.

Abel and Vezzosi \cite{AbelV14} formalized a simple type theory
extended with the later modality in Agda. They focus on operational
properties of the calculus and give a certified proof of strong
normalization. In future work, we plan to investigate the metatheory
and the dynamic behaviour of \GTT, for example by formalizing a  type
checker and a normalization algorithm.

\remove{
and weakening, and we constructed denotational semantics for it in type theory extended with sized types.
From this, we concluded that this theory is consistent and that guarded recursion can be defined via sized types.

There are several possible ways to improve on this result.
First of all, one could define a dependent version of this type theory extended with multiple clocks.
This can be done in two ways.
Either one extends the object theory with delayed substitutions \cite{bizjak2016guarded} or with ticks \cite{BahrGM17}.
Second of all, with the deep embedding in place, one can further study the meta-theory by defining a formalized type checking and a normalization algorithm.
}

Finally, we want to understand weather it is possible to reduce sized types to guarded
recursion.  This does not seems directly doable, since sizes are not uniquely employed with coinductive types as a
tool for easing the productivity check of corecursive
definitions, but they are also used as auxiliary arguments of inductive
types for facilitating the termination check of recursive
definitions. It seems that the existing type theories for guarded
recursion are not capable of capturing this dual aspect of sizes.
